Stochastic blockmodeling of relational event dynamics

### 2012-05-31 docs/intro.tex
Fix up intro example figure.
edited the intro (look for "PS" in the comments)

Statistical methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions \cite{Goldenberg2009}.
One unifying idea is that the nodes of these networks are differentiated by rates of interacting with other parts of the network [TODO] that is associated with some other quantity, e.g. a covariate such as age or sex, or perhaps shared [TODO].
Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez} are a class of statistical models for static network data that employ latent variables to model this heterogeneity by 1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of edges between nodes of each block.

If we are interested in studying the events in a network, there are other aspects of the dynamics in addition to the mean rate. 
For example, human conversation often has an increased propensity for reciprocated events.
Shared tendencies based on other observed or unobserved quantities is of interest (who shares this tendency).
[TODO: Example of a shared dynamic... perhaps sticking with the reciprocity bit.]
Figure \ref{fig:introexample} shows how an analysis using a blockmodels will be unable to learn about these differences.

@@

Borrowing from the intuition of stochastic blockmodels, we propose a continuous-time model of network-based event sequences where latent clusters of nodes share similar patterns of interaction.
Our approach employs a flexible framework for  specifing how the process depends on the previous history of events \cite{AalenOddO.2008, Butts2008}.
%extend continuous-time models from event history analysis to model sequences of network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}. % \cite{AalenOddO.2008} 
%These models allow one to specify how the process depends on the previous history of events. % TODO use 'flexible'?
In this way one may investigate theories about the underlying processes and make predictions about future data conditioned on the past.

@@

\begin{figure}
\centering
\includegraphics[scale=.8]{../figs/introexample/all}
\caption{Illustration of differentiation by rate of interaction versus by dynamics.  The left column shows an example where within-block communication is large, and there is a higher tendency for reciprocity within a block.  In the situation where the two groups are not differentiated by rate, as in the right column, a standard blockmodel is unable to distinguish between groups A and B.  The proposed method can, however, learn such groups by employing more flexible definitions of shared dynamics.}
\label{fig:example}
\end{figure}

>>>>

Statistical methods for analyzing network data are becoming increasingly useful for studying phenomena ranging from online social behavior to protein interactions \cite{Goldenberg2009}.
One unifying idea is that the nodes of these networks are differentiated by rates of interacting with other parts of the network [TODO], where this variation in rates may be associated with some unobserved quantity, e.g. a covariate such as age or sex, or perhaps shared [TODO].
Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez} are a class of statistical models for static network data that employ latent variables to model this heterogeneity by 1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of edges between nodes of each block. 
%PS: much of this past work is for static network data. Do we want to make the observation here that we are thinking about blockmodels for *rate* data here in our discussion and in our example in figure 1 - if we don't some readers/reviewers might get a little confused. For example we could say that instead of a binomial model on edges we can think of a Poisson process yielding counts over time, where we can still think about blockmodels. 
%PS: also, we introduce the term "rates" above without really defining it, and then go on to talk about "mean rate" below. So I think we definitely need a sentence somewhere above that makes it clear we are talking about edges that instantaneously appear and disappear over time at some rate, rather than static edges.

If we are interested in studying a network of events, there are other important aspects of the dynamics in addition to the mean rate. 
For example, human conversation often has an increased propensity for reciprocated events.
%PS: wording of next sentence is a little odd - not sure I understand what it means...
Shared tendencies based on other observed or unobserved quantities is of interest (who shares this tendency).
[TODO: Example of a shared dynamic... perhaps sticking with the reciprocity bit.]
Figure \ref{fig:introexample} shows a simple example of how traditional blockmodels are unable to make inferences these types of dynamic phenomena.

@@

Borrowing from the intuition of stochastic blockmodels, we propose a continuous-time model of network-based event sequences where latent clusters of nodes share similar patterns of interaction.
Our approach employs a flexible framework for  specifying how the process depends on the previous history of events \cite{AalenOddO.2008, Butts2008}.
%extend continuous-time models from event history analysis to model sequences of network-based event data \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}. % \cite{AalenOddO.2008} 
%These models allow one to specify how the process depends on the previous history of events. % TODO use 'flexible'?
In this way one can investigate theories about the underlying processes and make predictions about future data conditioned on the past.

@@

\begin{SCfigure}
\centering
\includegraphics[scale=.45]{../figs/introexample/all}
\caption{Illustration of differentiation by rate of interaction versus by dynamics.  The left column shows an example where within-block communication is large, and there is a higher tendency for reciprocity within a block.  In the situation where the two groups are not differentiated by rate, as in the right column, a standard blockmodel is unable to distinguish between groups A and B.  The proposed method can, however, learn such groups by employing more flexible definitions of shared dynamics.}
\label{fig:example}
\end{SCfigure}
